---
title: "Minimum sample size for SSD HC~x~ estimation"
subtitle: "Small-N bias, confidence-interval width, and coverage from the `study_small_hpc` simulation"
date: today
format:
  html:
    toc: true
    fig-align: center
    df-print: kable
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)

# Reads the artifacts written by make_report_outputs.R (run that first):
#   ../output/study_small_report_figure.png
#   ../output/study_small_summary_table.csv
fig_path <- "../output/study_small_report_figure.png"
tbl_path <- "../output/study_small_summary_table.csv"
stopifnot(file.exists(fig_path), file.exists(tbl_path))

summary_table <- read.csv(tbl_path)
```

## Study design

For each of 41 ANZG/CCME example datasets (34 ANZG, 7 CCME — all such datasets in
`ssddata`), the top-AICc-weighted distribution was treated as the
data-generating "truth". For every dataset, 500 samples of each size
$N \in \{5, 6, 7, 8, 10, 16, 26\}$ were drawn and re-fit with the model-averaged
("multi") method, with 95% confidence intervals from a weighted-sample
bootstrap. Three quantities are reported against $N$, for four HC~x~ proportions
(HC1, HC5, HC10, HC20):

- **Relative bias** — $(\hat{}\text{HC}_x - \text{HC}_x)/\text{HC}_x$
- **Relative CI width** — $(\text{ucl} - \text{lcl})/\text{HC}_x$
- **Coverage** — the fraction of simulations whose CI contained the true HC~x~

## Computation and run environment

The simulation was run on the AIMS HPC under SLURM, orchestrated with the
[`targets`](https://docs.ropensci.org/targets/) pipeline framework and a
[`crew.cluster`](https://wlandau.github.io/crew.cluster/) SLURM controller: each
work "shard" is submitted as its own SLURM job, while the driver process runs on
a login node. Sharding is coarsened so that each of the sampling,
distribution-fitting, and HC~x~ steps produces one shard per simulation (all 41
datasets bundled together) — trading some parallelism for longer-lived jobs that
suit the scheduler.

Worker jobs requested 1 CPU and 8 GB RAM on the `cpuq` partition, with a 12-hour
wall limit and a pool of up to 120 concurrent workers. The 8 GB was necessary:
at 2 GB the HC~x~ workers were killed for exceeding memory. Fitting used R 4.4.1
with `ssdsims`/`ssdtools`, and results were written as local Parquet. The figure
and tables in this report were produced afterwards on a WSL workstation.

## Figure

```{r figure}
#| fig-cap: "HC~x~ estimation versus sample size, faceted by HC proportion. Red boxes mark N ≤ 6. Relative bias is on a symmetric-log axis (it is signed, spanning −1 to ~10^8^); relative CI width is on a log10 axis; coverage is linear, with the dashed line at the nominal 0.95."
include_graphics(fig_path)
```

## Interpretation

**Coverage collapses at N = 5.** Coverage of the true HC~x~ falls to 0.58 (HC1)
through 0.78 (HC20) at $N = 5$ — well below the nominal 0.95 — then jumps to
roughly 0.87–0.89 at $N = 6$ and creeps up to about 0.93–0.94 by $N = 26$.
$N = 5$ is clearly the deficient case; $N = 6$ already recovers most of the lost
coverage.

**Confidence intervals are enormous at small N.** The median relative CI width
is about 41× the true HC1 value at $N = 5$, falling to roughly 3.9× at
$N = 26$. Even at $N = 26$ the low-proportion intervals remain wide — the honest
signal that a small sample cannot pin down an extreme lower-tail quantile.

**Bias is dominated by a heavy positive tail.** Median bias is modest, but the
upper tail reaches order 10^8^ at the smallest N and lowest proportions (visible
as the long whiskers in the top row). One pattern worth noting: for **HC1**, the
median bias is **+2.05 at $N = 5$**, flips **negative (−0.24 to −0.32) at
$N = 6$–10**, then returns toward zero — a non-monotonic trend that reflects the
small-N model-averaging / fallback behaviour rather than a smooth decline.

**Takeaway.** Across all three metrics, $N = 5$ is a distinct failure regime and
$N = 6$ is the point at which behaviour becomes substantially more reliable,
consistent with $N = 5$ being the boundary this study set out to probe.

## Summary table

```{r table}
#| tbl-cap: "Coverage, median relative bias (with interquartile range), and median relative CI width by HC proportion and sample size. Metric values are rounded to 3 significant figures."
display <- summary_table |>
  mutate(
    HC = paste0("HC", 100 * proportion),
    across(c(coverage, bias_median, bias_q25, bias_q75, width_median),
      \(x) signif(x, 3))
  ) |>
  select(HC, N = nrow, Sims = n_sims, Coverage = coverage,
    `Median bias` = bias_median, `Bias Q25` = bias_q25,
    `Bias Q75` = bias_q75, `Median CI width` = width_median)

kable(display, align = "lrrrrrrr")
```

## Source-distribution AICc weights

Each dataset's simulations were generated from its own top-AICc-weighted BCANZ
distribution — its "source" distribution. How cleanly a dataset represents a
single distribution is itself informative, and it determines whether we can
isolate per-distribution behaviour (as in Figure 4 of the original study). The
per-dataset weights are extracted from the cached fits into
`dataset_aicc_weights.csv`.

```{r weights}
#| tbl-cap: "For each BCANZ distribution: how many of the 41 datasets it is the top-weighted (source) distribution for, and the single dataset where it attains its highest AICc weight."
w <- read.csv("dataset_aicc_weights.csv")
dists_bcanz <- c("gamma", "lgumbel", "llogis", "lnorm", "lnorm_lnorm", "weibull")

wt_summary <- do.call(rbind, lapply(dists_bcanz, function(d) {
  i <- which.max(w[[d]])
  data.frame(
    Distribution = d,
    `Datasets sourced` = sum(w$source_dist == d),
    `Most representative dataset` = w$dataset[i],
    `Max AICc weight` = round(w[[d]][i], 3),
    check.names = FALSE
  )
}))
kable(wt_summary, align = "lrlr")

mix_i <- which.max(w$lnorm_lnorm)
mix_ds <- w$dataset[mix_i]; mix_wt <- w$lnorm_lnorm[mix_i]
uni <- setdiff(dists_bcanz, "lnorm_lnorm")
uni_max <- vapply(uni, function(d) max(w[[d]]), numeric(1))
```

Across the 41 ssddata datasets, only the **lnorm_lnorm mixture has a near-pure
representative** (`` `r mix_ds` ``, weight `r round(mix_wt, 2)`); for the five
unimodal distributions the best any single dataset reaches is only
`r round(min(uni_max), 2)`–`r round(max(uni_max), 2)`, so no ssddata dataset is
unambiguously one distribution. This is why the original study drew on the much
larger 353-dataset EnviroTox corpus — to find clean, high-weight representatives
of each source distribution. The full 41 × 6 weight matrix is in
`dataset_aicc_weights.csv`.

## Refitted vs. true distribution weights

The original study's Figure 4 simulated from each dataset's single *true* source
distribution and tracked how the AICc weight of that distribution grew with N.
Our study simulates from the model-averaged *multi* fit (all six components,
weighted by their real-data AICc weights), so there is no single "true" source
per dataset — instead each dataset has a *vector* of true weights. The natural
analogue is therefore to ask: **as N grows, do the re-fitted AICc weights recover
the original (real-data) weights?** Below, each point is one dataset at one N:
its observed mean re-fitted weight for a distribution (averaged over the 500
simulations) against that distribution's true weight from the real-data fit.

These weights are a fit-step property not present in `summary.parquet`; they are
extracted from the 55 GB of `fit` shards on the HPC by decoding each shard's
serialised `fitdists` and reading `ssdtools::glance(wt = TRUE)`. That extraction
(`extract_fit_weights.R`) is run as a SLURM job and its small result CSV pulled
to WSL — the 55 GB never leaves the cluster:

```sh
# on the HPC (from study_small_hpc/):
module load R/4.4.1 slurm
sbatch extract_fit_weights.sbatch      # ~20-25 min on 8 cores -> fit_weights_observed.csv

# then from WSL, pull the compact result (not the shards):
REMOTE=/export/home/q-z/rfisher/ssdtools_set_min_n/ssdsims_pipeline/study_small_hpc
rsync -az "rfisher@hpc-l001.aims.gov.au:$REMOTE/fit_weights_observed.csv" ./
```

The raw extract (`fit_weights_observed.csv`, ~20 MB, one row per fit) is
regenerable and git-ignored; it is aggregated once to the committed
`fit_weights_summary.csv` (mean weight per dataset × N × distribution) that the
figure below reads:

```r
library(dplyr); library(tidyr)
dists <- c("gamma", "lgumbel", "llogis", "lnorm", "lnorm_lnorm", "weibull")
read.csv("fit_weights_observed.csv") |>
  pivot_longer(all_of(dists), names_to = "distribution", values_to = "weight") |>
  group_by(dataset, nrow, distribution) |>
  summarise(mean_weight = mean(weight), sd_weight = sd(weight), n_sims = n(),
    .groups = "drop") |>
  write.csv("fit_weights_summary.csv", row.names = FALSE)
```

```{r weights-convergence}
#| fig-cap: "Observed mean re-fitted AICc weight (averaged over 500 simulations per dataset and N) versus the true weight from the real-data fit, for each of the six BCANZ distributions. One point per dataset per N; the dashed line is y = x (observed = true)."
#| fig-width: 9
#| fig-height: 6
dists_bcanz <- c("gamma", "lgumbel", "llogis", "lnorm", "lnorm_lnorm", "weibull")

obs_long <- read.csv("fit_weights_summary.csv") |>
  transmute(dataset, nrow, distribution, observed = mean_weight)

true_long <- read.csv("dataset_aicc_weights.csv") |>
  pivot_longer(all_of(dists_bcanz), names_to = "distribution", values_to = "true") |>
  select(dataset, distribution, true)

wc <- left_join(obs_long, true_long, by = c("dataset", "distribution")) |>
  mutate(
    distribution = factor(distribution, dists_bcanz),
    N = factor(nrow, levels = sort(unique(nrow)))
  )

# convergence summaries referenced in the text below
cor_by_n <- wc |> group_by(N) |> summarise(r = cor(observed, true), .groups = "drop")
mix_by_n <- wc |> filter(distribution == "lnorm_lnorm") |> group_by(N) |>
  summarise(m = mean(observed), .groups = "drop")
gv <- function(df, col, n) df[[col]][df$N == n]
cor5 <- gv(cor_by_n, "r", "5"); cor6 <- gv(cor_by_n, "r", "6"); cor26 <- gv(cor_by_n, "r", "26")
mix5 <- gv(mix_by_n, "m", "5"); mix6 <- gv(mix_by_n, "m", "6")

ggplot(wc, aes(true, observed, colour = N)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(size = 0.8, alpha = 0.7) +
  facet_wrap(~distribution) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_colour_viridis_d(option = "C", end = 0.9) +
  labs(x = "True AICc weight (real-data fit)",
    y = "Observed mean re-fitted AICc weight") +
  theme_bw(base_size = 10)
```

At **N = 5 the lnorm_lnorm mixture captures almost all of the AICc weight** —
mean `r round(mix5, 2)` across *every* dataset, regardless of its true mixture
weight — which forces the five unimodal distributions to ~0 and makes the
re-fitted weights *anti-correlated* with the truth (r = `r round(cor5, 2)`). In
the figure this is the dark N = 5 cluster pinned near 1 in the lnorm_lnorm panel
and on the zero floor of every unimodal panel. With only five points the flexible
five-parameter mixture spuriously "wins".

From **N = 6 the mixture is correctly penalised** to essentially zero (mean
`r round(mix6, 3)`) — except for the one genuinely bimodal dataset
(`anzg_iron_marine`), which earns high mixture weight only at larger N — and the
unimodal weights begin to track their true values: the correlation climbs from
`r round(cor6, 2)` (N = 6) to `r round(cor26, 2)` (N = 26).

Recovery is directional but **incomplete**: even at N = 26 the points sit below
the y = x line, because model averaging over six candidates spreads the weight
and the real-data weights are themselves diffuse (only one dataset exceeds 0.9,
above). The message mirrors both the coverage result and the original study's
Figure 4B: the mixture should be preferred only with genuine bimodality *and*
adequate N, and N = 5 is again a distinct failure regime — here, spurious
mixture dominance.

## Run diagnostics

Build times are computed from a per-shard extract of the run's `targets`
metadata store (`run_diagnostics.csv`, committed alongside this report). See
[Reproducing these diagnostics] for exactly how that file was pulled from the
HPC.

```{r diagnostics}
#| tbl-cap: "Per-step build diagnostics computed from the run's targets metadata."
diag <- read.csv("run_diagnostics.csv")

# Reference specs for the WSL workstation this report compares against, and the
# run's SLURM sizing (the hc workers failed at 2 GB and ran at 8 GB).
wsl_cores   <- 22L
wsl_ram_gb  <- 31
hc_shard_gb <- 8
hpc_workers <- 120L

step_order <- c("sample", "fit", "hc", "summary", "barrier")
fmt_time <- function(h) ifelse(h < 1 / 60, "<0.1 min",
  ifelse(h < 1, paste0(round(h * 60, 1), " min"), paste0(round(h, 1), " h")))

by_step <- diag |>
  dplyr::group_by(step) |>
  dplyr::summarise(
    shards  = dplyr::n(),
    total_h = sum(seconds, na.rm = TRUE) / 3600,
    mean_h  = mean(seconds, na.rm = TRUE) / 3600,
    max_h   = suppressWarnings(max(seconds, na.rm = TRUE)) / 3600,
    gb      = sum(bytes, na.rm = TRUE) / 1e9,
    .groups = "drop"
  ) |>
  dplyr::mutate(step = factor(step, step_order)) |>
  dplyr::arrange(step)

# Scalars reused in the prose/estimates below.
total_h    <- sum(diag$seconds, na.rm = TRUE) / 3600
total_gb   <- sum(diag$bytes, na.rm = TRUE) / 1e9
n_targets  <- nrow(diag)
i_long     <- which.max(diag$seconds)
long_name  <- diag$name[i_long]
long_h     <- diag$seconds[i_long] / 3600
get <- function(s, col) by_step[[col]][by_step$step == s]
hc_h <- get("hc", "total_h"); hc_n <- get("hc", "shards")
fit_h <- get("fit", "total_h"); fit_gb <- get("fit", "gb")

serial_days <- total_h / 24
ideal_h     <- total_h / wsl_cores
ram_conc    <- floor(wsl_ram_gb / hc_shard_gb) # hc shards that fit in RAM at once
ram_days    <- total_h / ram_conc / 24
hpc_ideal_h <- total_h / hpc_workers

by_step |>
  dplyr::transmute(
    Step = as.character(step),
    Shards = format(shards, big.mark = ","),
    `Total compute` = fmt_time(total_h),
    `Mean / shard` = fmt_time(mean_h),
    `Longest shard` = fmt_time(max_h),
    Output = paste0(round(gb, 2), " GB")
  ) |>
  kable(align = "lrrrrr")
```

The pipeline comprised **`r format(n_targets, big.mark = ",")` targets**. The
HC~x~ bootstrap dominates: `r hc_n` shards, ~`r round(hc_h)` core-hours in total,
the longest (`` `r long_name` ``) taking `r round(long_h, 1)` h. Distribution
fitting adds only ~`r round(fit_h)` core-hours but ~`r round(fit_gb)` GB of the
`r round(total_gb, 1)` GB of output (fitted `fitdists` objects are large); the
sampling and summary steps are negligible in both time and size. The **total
serial compute is ~`r format(round(total_h), big.mark = ",")` core-hours
(≈ `r round(serial_days)` core-days)**.

Spread across the pool of up to `r hpc_workers` SLURM workers, that is
~`r round(hpc_ideal_h, 1)` h of *ideal* wall time — consistent with the ~7 h
observed for the final (post-fix) run.

### What this would cost on the WSL workstation

The same workload on the WSL machine (`r wsl_cores` cores, `r wsl_ram_gb` GB
RAM), assuming comparable per-core speed:

| Scenario | Estimated wall time |
|---|---|
| Single core (serial) | ~`r format(round(total_h), big.mark = ",")` h ≈ **`r round(serial_days)` days** |
| All `r wsl_cores` cores, *if* memory allowed | ~`r round(ideal_h)` h ≈ **`r round(ideal_h / 24, 1)` days** |
| Realistic (memory-bound) | ~`r round(ram_days)` days (**~`r round(ram_days / 7, 1)` weeks**) |

The all-cores figure is not achievable here: each HC~x~ shard needs several GB of
RAM, so `r wsl_ram_gb` GB caps concurrency at ~`r ram_conc` shards, not
`r wsl_cores` — the run would be bound by memory, not cores. (If the true peak is
below the 8 GB requested, more shards fit and the figure shrinks toward a week.)
In short, the HPC's parallelism *and* per-node memory turned a **multi-week**
single-workstation job into roughly **eight hours** of wall time.

### Reproducing these diagnostics

`run_diagnostics.csv` is a per-shard extract of the HPC run's `targets` metadata.
To regenerate it, pull the metadata file from the HPC (read-only; from WSL):

```sh
REMOTE=/export/home/q-z/rfisher/ssdtools_set_min_n/ssdsims_pipeline/study_small_hpc
rsync -az "rfisher@hpc-l001.aims.gov.au:$REMOTE/_targets/meta/meta" ./hpc_meta.tsv
```

then parse the pipe-delimited metadata into the committed CSV:

```r
m <- read.delim("hpc_meta.tsv", sep = "|")
m$seconds <- as.numeric(m$seconds)
m$bytes   <- as.numeric(m$bytes)
m$step <- ifelse(grepl("^sample_step", m$name), "sample",
          ifelse(grepl("^fit_step", m$name), "fit",
          ifelse(grepl("^hc_step", m$name), "hc",
          ifelse(grepl("^summary", m$name), "summary", "barrier"))))
write.csv(m[, c("name", "step", "seconds", "bytes")],
  "run_diagnostics.csv", row.names = FALSE)
```

---

*Generated from `study_small_hpc` outputs. Rebuild the figure and table with
`Rscript make_report_outputs.R` before re-rendering. Run-diagnostic figures are
from the `targets` metadata of the completed HPC run.*
